---
title: "New year new me form 2019-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(matrixStats)
```

First we shall load and merge all our data.

```{r}
prem2019_20 <- read.csv("prem2019-20.csv")
prem2019_20<-data.frame(Date=prem2019_20$Date,HomeTeam=prem2019_20$HomeTeam,AwayTeam=prem2019_20$AwayTeam,FTHG=prem2019_20$FTHG,FTAG=prem2019_20$FTAG)

fulldf<-(prem2019_20)
fulldf
```

Next we find all the teams which have games included in our data set as well as defining $tn$ as the total number of teams.

```{r}
fullteams<-function(dt){
  gamenumber<-length(dt$HomeTeam)
  teams <- vector()
  for (i in 1:gamenumber){
    if (dt$HomeTeam[i] %in% teams==FALSE){
      teams <- append(teams, dt$HomeTeam[i])
      }
  }
  return(teams)
  }
fullteamlist<-sort(fullteams(fulldf))
tn<-length(fullteamlist)
```

Now we define the vector monthd. This contains the number of days in each month.

```{r}
monthd<-c(31,28,31,30,31,30,31,31,30,31,30,31)
```

As our weighting function uses $t$ in units of days back we need to define a function that, using the date a match took place on and the date on which we are predicting each team's strength, can calculate how many days back a match took place.

```{r}
datetodays<-function(date){
  year<-as.numeric(substr(date,7,10))
  month<-as.numeric(substr(date,4,5))
  day<-as.numeric(substr(date,1,2))
  yeardays<-365*(year-1)
  monthdays<-0
  for (i in 1:month){
    monthdays<-monthdays+monthd[i]
  }
  if (year==2020){
    if (month>1){
      monthdays<-monthdays+1
    }
  }
  days<-day+monthdays+yeardays
  return(days)
  }

```

We now convert our datafram so it only involves the relevant information. That involves converting the date each match took place on into how many days it took place before the data set. This also involves defining the result of each match by comparing how many goals each team scored in each match.

```{r}
converted<-function(dt){
  teams<-fullteams(dt)
  n<-length(teams)
  gn<-length(dt$HomeTeam)
  finaldate<-dt$Date[gn]
  finaldays<-datetodays(finaldate)
  TeamWins<-data.frame(Date=character(0),HomeTeam=character(0),AwayTeam=character(0),Result=integer(0),DaysBack=integer(0))
  for (i in 1:gn){
    Date<-dt$Date[i]
    HomeTeam<-dt$HomeTeam[i]
    AwayTeam<-dt$AwayTeam[i]
    if (dt$FTHG[i]>dt$FTAG[i]){
          Result<-1
          }
        if (dt$FTHG[i]<dt$FTAG[i]){
          Result<-(-1)
          }
        if (dt$FTHG[i]==dt$FTAG[i]){
          Result<-0
        }
    DaysBack<-finaldays-datetodays(dt$Date[i])
    gameresult<-data.frame(Date,HomeTeam, AwayTeam, Result, DaysBack)
    TeamWins<- rbind(TeamWins, gameresult)
    }
  return(TeamWins)
}

fulldf<-converted(fulldf)
fulldf
```

We now code our log-likelihood function that takes in account home advantage and draws. This is $$\ell(\pi,\gamma,\theta)=\sum_{i}\sum_{j}[(a_{ij}+v_{ij})\ln{(\gamma\pi_i)}-(a_{ij}+v_{ij})\ln{(\gamma\pi_i+\theta\pi_j)}+(b_{ij}+v_{ij})\ln{(\pi_j)}-(b_{ij}+v_{ij})\ln{(\pi_j+\theta\gamma\pi_i)}+v_{ij}\ln{(\theta^2-1)}]$$. We define the strength of the last team to be 0.1. We use $\pi_i=\exp{(b_i)}$ to ensure all our values are positive and logSumExp for its computational advantages. We then use the optim function to find our maximum likelihood estimators. We then run this function using a $\epsilon=0.0057$ as this has previously been found to be the optimal value of $\epsilon$.

```{r}
MLE<-function(eps,df){
  dc<-eps
  gn<-length(df$HomeTeam)
  WDHLL3<-function(theta){
    theta[tn+2]<-theta[tn+1]
    theta[tn+1]<-theta[tn]
    theta[tn]<-log(0.1)
    D<-theta[tn+1]
    H<-theta[tn+2]
    ans<-0
    for (i in 1:gn){
      Home<-df$HomeTeam[i]
      Away<-df$AwayTeam[i]
      Result<-df$Result[i]
      Days<-df$DaysBack[i]
      I<-theta[which(fullteamlist==Home)]
      J<-theta[which(fullteamlist==Away)]
      decay<-exp(-(dc*Days))
      if (Result==1){
        parta<-(H+I)-(logSumExp(c(H+I,D+J)))
      }
      if (Result==-1){
        parta<-(J)-(logSumExp(c(J,D+H+I)))
      }
      if (Result==0){
        parta<-(log((exp(2*D))-1))+(H+I+J)-(logSumExp(c(H+I,D+J)))-(logSumExp(c(J,D+H+I)))
      }
      ans<-ans+(decay*parta)
      
    }
    return(ans)

  }
  qq<-c(rep(1,tn-1),1.5,1.8)
  lowlim<-c(rep(-100,tn-1),1e-5,1e-5)
  strengthDH2<-optim(theta <- qq, WDHLL3, method="L-BFGS-B",lower=lowlim,control=list(fnscale=-1,maxit=1000),hessian=TRUE)
  return(strengthDH2)
}
  
withoutform<-MLE(0,fulldf)
withform<-MLE(0.0022,fulldf)
```

Finally we rank the estimated strengths of each team as well as stating the estimated draw and home advantage parameters. First we do this when form is not taken into account. 

```{r}
paramnames<-fullteamlist
paramnames<-append(paramnames,c("theta","gamma"))

inv_fisher_info_fans<-solve(-withoutform$hessian)
prop_sigma_no_form<-sqrt(diag(inv_fisher_info_fans))
prop_sigma_no_form[tn+2]<-prop_sigma_no_form[tn+1]
prop_sigma_no_form[tn+1]<-prop_sigma_no_form[tn]
prop_sigma_no_form[tn]<-"N/A"

noformmle<-exp(withoutform$par)
noformmle[tn+2]<-noformmle[tn+1]
noformmle[tn+1]<-noformmle[tn]
noformmle[tn]<-0.1

noform.df<-data.frame("Teams"=paramnames,"MLEs"=noformmle,"S.E"=prop_sigma_no_form)
noform.df[order(noform.df$MLEs),]
```

Next we do this when form is taken into account.

```{r}
inv_fisher_info_fans<-solve(-withform$hessian)
prop_sigma_form<-sqrt(diag(inv_fisher_info_fans))
prop_sigma_form[tn+2]<-prop_sigma_form[tn+1]
prop_sigma_form[tn+1]<-prop_sigma_form[tn]
prop_sigma_form[tn]<-"N/A"

formmle<-exp(withform$par)
formmle[tn+2]<-formmle[tn+1]
formmle[tn+1]<-formmle[tn]
formmle[tn]<-0.1

form.df<-data.frame(Teams=paramnames,MLEs=formmle,S.E=prop_sigma_form)
form.df[order(form.df$MLEs),]
```



