---
title: "new year new me home advantage analysis with grad"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(matrixStats)
```

First we shall load and merge all our data.

```{r}
prem2020_21 <- read.csv("prem2020-21 more recent.csv")
prem2020_21<-data.frame(Date=prem2020_21$Date,HomeTeam=prem2020_21$HomeTeam,AwayTeam=prem2020_21$AwayTeam,FTHG=prem2020_21$FTHG,FTAG=prem2020_21$FTAG)

prem2019_20 <- read.csv("prem2019-20.csv")
prem2019_20<-data.frame(Date=prem2019_20$Date,HomeTeam=prem2019_20$HomeTeam,AwayTeam=prem2019_20$AwayTeam,FTHG=prem2019_20$FTHG,FTAG=prem2019_20$FTAG)

prem2018_19 <- read.csv("prem2018-19.csv")
prem2018_19 <- data.frame(Date=prem2018_19$Date,HomeTeam=prem2018_19$HomeTeam,AwayTeam=prem2018_19$AwayTeam,FTHG=prem2018_19$FTHG,FTAG=prem2018_19$FTAG)

prem2017_18 <- read.csv("prem2017-18.csv")
prem2017_18 <- data.frame(Date=prem2017_18$Date,HomeTeam=prem2017_18$HomeTeam,AwayTeam=prem2017_18$AwayTeam,FTHG=prem2017_18$FTHG,FTAG=prem2017_18$FTAG)

prem2016_17 <- read.csv("prem2016-17.csv")
prem2016_17 <- data.frame(Date=prem2016_17$Date,HomeTeam=prem2016_17$HomeTeam,AwayTeam=prem2016_17$AwayTeam,FTHG=prem2016_17$FTHG,FTAG=prem2016_17$FTAG)

prem2015_16 <- read.csv("prem2015-16.csv")
prem2015_16 <- data.frame(Date=prem2015_16$Date,HomeTeam=prem2015_16$HomeTeam,AwayTeam=prem2015_16$AwayTeam,FTHG=prem2015_16$FTHG,FTAG=prem2015_16$FTAG)

fulldf<-rbind(prem2015_16,prem2016_17,prem2017_18,prem2018_19,prem2019_20,prem2020_21)
fulldf
```

We now code the functions "datestonumbersbefore" and "datestonumbersafter" that will split the data into two seperate data frames. The first including all the matches before a specific date and the second including all matches after a specific date. We then run both these functions using the date 16/06/2020 as this is when fans stopped being allowed to attend Premier-League matches.

```{r}
datestonumbersbefore<-function(dt,endday,endmonth,endyear){
  gn<-length(dt$Date)
  for (i in 1:length(dt$Date)){
    year<-as.numeric(substr(dt$Date[i],7,10))
    if (year>endyear){
      dt <- dt[-(i:gn), ]
      break
    }
    if (year==endyear){
      month<-as.numeric(substr(dt$Date[i],4,5))
      if (month>endmonth){
        dt <- dt[-(i:gn), ]
        break
      }
      if (month==endmonth){
        day<-as.numeric(substr(dt$Date[i],1,2))
        if (day>endday){
          dt <- dt[-(i:gn), ]
          break
        }
      }
    }
    }
  return(dt)
}

datestonumbersafter<-function(dt,startday,startmonth,startyear){
  gn<-length(dt$Date)
  for (i in (gn:1)){
    year<-as.numeric(substr(dt$Date[i],7,10))
    if (year<startyear){
      dt <- dt[-(1:i), ]
      break
    }
    if (year==startyear){
      month<-as.numeric(substr(dt$Date[i],4,5))
      if (month<startmonth){
        dt <- dt[-(1:i), ]
        break
      }
      if (month==startmonth){
        day<-as.numeric(substr(dt$Date[i],1,2))
        if (day<=startday){
          dt <- dt[-(1:i), ]
          break
        }
      }
    }
  }
  return(dt)
}


fulldfnofans<-datestonumbersafter(fulldf,16,6,2020)
fulldfnofans
fulldfwithfans<-datestonumbersbefore(fulldf,16,6,2020)
fulldfwithfans
```

Next define a function that finds all the teams which have games included in our data set.

```{r}
fullteams<-function(dt){
  gamenumber<-length(dt$HomeTeam)
  teams <- vector()
  for (i in 1:gamenumber){
    if (dt$HomeTeam[i] %in% teams==FALSE){
      teams <- append(teams, dt$HomeTeam[i])
      }
  }
  return(teams)
}
```

Now we define the vector monthd. This contains the number of days in each month.

```{r}
monthd<-c(31,28,31,30,31,30,31,31,30,31,30,31)

```

As our weighting function uses $t$ in units of days back we need to define a function that, using the date a match took place on and the date on which we are predicting each team's 
strength, can calculate how many days back a match took place.

```{r}
datetodays<-function(date){
  year<-as.numeric(substr(date,7,10))
  month<-as.numeric(substr(date,4,5))
  day<-as.numeric(substr(date,1,2))
  yeardays<-365*(year-1)
  monthdays<-0
  for (i in 1:month){
    monthdays<-monthdays+monthd[i]
  }
  if (year==2020){
    if (month>1){
      monthdays<-monthdays+1
    }
  }
  days<-day+monthdays+yeardays
  return(days)
  }

```

We now convert our dataframe so it only involves the relevant information. That involves converting the date each match took place on into how many days it took place before the data set. This also involves defining the result of each match by comparing how many goals each team scored in each match.

```{r}
converted<-function(dt){
  teams<-fullteams(dt)
  tn<-length(teams) # the total number of teams in the data set
  gn<-length(dt$HomeTeam) # the total number of games
  finaldate<-dt$Date[gn]
  finaldays<-datetodays(finaldate)
  TeamWins<-data.frame(Date=character(0),HomeTeam=character(0),AwayTeam=character(0),Result=integer(0),DaysBack=integer(0))
  for (i in 1:gn){
    Date<-dt$Date[i]
    HomeTeam<-dt$HomeTeam[i]
    AwayTeam<-dt$AwayTeam[i]
    if (dt$FTHG[i]>dt$FTAG[i]){
          Result<-1
          }
        if (dt$FTHG[i]<dt$FTAG[i]){
          Result<-(-1)
          }
        if (dt$FTHG[i]==dt$FTAG[i]){
          Result<-0
        }
    DaysBack<-finaldays-datetodays(dt$Date[i])
    gameresult<-data.frame(Date,HomeTeam, AwayTeam, Result, DaysBack)
    TeamWins<- rbind(TeamWins, gameresult)
    }
  return(TeamWins)
}

fulldfnofans<-converted(fulldfnofans)
fulldfnofans
fulldfwithfans<-converted(fulldfwithfans)
fulldfwithfans

```
We now code our log-likelihood function that takes in account home advantage and draws. This is $$\ell(\pi,\gamma,\theta)=\sum_{i}\sum_{j}[(a_{ij}+v_{ij})\ln{(\gamma\pi_i)}-(a_{ij}+v_{ij})\ln{(\gamma\pi_i+\theta\pi_j)}+(b_{ij}+v_{ij})\ln{(\pi_j)}-(b_{ij}+v_{ij})\ln{(\pi_j+\theta\gamma\pi_i)}+v_{ij}\ln{(\theta^2-1)}]$$. We define the strength of the last team to be 0.1. We use $\pi_i=\exp{(b_i)}$ to ensure all our values are positive and logSumExp for its computational advantages. We then use the optim function to find our maximum likelihood estimators for each set of data. We then run this function using a $\epsilon=0$ as there is no need to worry about form when comparing home advantage. 

```{r}
MLE<-function(eps,df){
  dc<-eps
  fullteamlist<-sort(fullteams(df))
  tn<-length(fullteamlist)
  gn<-length(df$HomeTeam)
  WDHLL3<-function(theta){
    D<-theta[tn]
    H<-theta[tn+1]
    theta[tn]<-log(0.1)
    ans<-0
    for (i in 1:gn){
      Home<-df$HomeTeam[i]
      Away<-df$AwayTeam[i]
      Result<-df$Result[i]
      Days<-df$DaysBack[i]
      I<-theta[which(fullteamlist==Home)]
      J<-theta[which(fullteamlist==Away)]
      decay<-exp(-(dc*Days))
      if (Result==1){
        parta<-(H+I)-(logSumExp(c(H+I,D+J)))
      }
      if (Result==-1){
        parta<-(J)-(logSumExp(c(J,D+H+I)))
      }
      if (Result==0){
        parta<-(log((exp(2*D))-1))+(H+I+J)-(logSumExp(c(H+I,D+J)))-(logSumExp(c(J,D+H+I)))
      }
      ans<-ans+(decay*parta)
    }
    return(ans)

  }
  
  qq<-c(rep(1,tn-1),1.5,1.8)
  lowlim<-c(rep(-100,tn-1),1e-5,1e-5)
  strengthDH2<-optim(theta <- qq, WDHLL3, method="L-BFGS-B",lower=lowlim,control=list(fnscale=-1),hessian=TRUE)
  return(strengthDH2)
}

withfans<-MLE(0,fulldfwithfans)
nofans<-MLE(0,fulldfnofans)

```

We then find the MLE for the difference in home parameters as well as the corresponding standard-error. We then use these values to define a $95\%$ confidence-interval for the home advantage with fans $\gamma_{1}$ subtracted by the home advantage without fans $\gamma_{2}$. If the value for the lower limit of the confidence interval is less than or equal to zero we say there is not significant evidence to reject the null hypothesis that the home advantage parameter has remained the same and hence we accept the null hypothesis. If the lower limit of the confidence interval is positive we have significant evidence to reject that the null hypothesis and thus accept the alternative hypothesis that home advantage has decreased. We then present our findings.

```{r}
inv_fisher_info_fans<-solve(-withfans$hessian)
prop_sigma_fans<-sqrt(diag(inv_fisher_info_fans))
n_param_fans<-length(withfans$par)
inv_fisher_info_no_fans<-solve(-nofans$hessian)
prop_sigma_no_fans<-sqrt(diag(inv_fisher_info_no_fans))
n_param_no_fans<-length(nofans$par)

standarderrors<-c(prop_sigma_fans[n_param_fans],prop_sigma_no_fans[n_param_no_fans])
print(standarderrors)
estimdiff<-(withfans$par[n_param_fans])-(nofans$par[n_param_no_fans])
total_var<-standarderrors[1]+standarderrors[2]

upper<-estimdiff+1.96*total_var
lower<-estimdiff-1.96*total_var
interval<-(data.frame("Estimated Difference"= estimdiff, "upper"=upper, "lower"=lower))
interval
```
Hence, as the lower limit of the 95% confidence-interval for the estimate of the the distance in home advantage parameters is more than zero, we have significant evidence to reject the null-hypothesis and accept the alternative hypothesis. Hence we deduce that home advantage has decreased.











