---
title: "new year new me analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(matrixStats)
```
First we shall load and merge all our data.
```{r}
prem2020_21 <- read.csv("prem2020-21.csv")
prem2020_21_ods<- prem2020_21
prem2020_21<-data.frame(Date=prem2020_21$Date,HomeTeam=prem2020_21$HomeTeam,AwayTeam=prem2020_21$AwayTeam,FTHG=prem2020_21$FTHG,FTAG=prem2020_21$FTAG)

prem2019_20 <- read.csv("prem2019-20.csv")
prem2019_20<-data.frame(Date=prem2019_20$Date,HomeTeam=prem2019_20$HomeTeam,AwayTeam=prem2019_20$AwayTeam,FTHG=prem2019_20$FTHG,FTAG=prem2019_20$FTAG)

prem2018_19 <- read.csv("prem2018-19.csv")
prem2018_19 <- data.frame(Date=prem2018_19$Date,HomeTeam=prem2018_19$HomeTeam,AwayTeam=prem2018_19$AwayTeam,FTHG=prem2018_19$FTHG,FTAG=prem2018_19$FTAG)

prem2017_18 <- read.csv("prem2017-18.csv")
prem2017_18 <- data.frame(Date=prem2017_18$Date,HomeTeam=prem2017_18$HomeTeam,AwayTeam=prem2017_18$AwayTeam,FTHG=prem2017_18$FTHG,FTAG=prem2017_18$FTAG)

prem2016_17 <- read.csv("prem2016-17.csv")
prem2016_17 <- data.frame(Date=prem2016_17$Date,HomeTeam=prem2016_17$HomeTeam,AwayTeam=prem2016_17$AwayTeam,FTHG=prem2016_17$FTHG,FTAG=prem2016_17$FTAG)

prem2015_16 <- read.csv("prem2015-16.csv")
prem2015_16 <- data.frame(Date=prem2015_16$Date,HomeTeam=prem2015_16$HomeTeam,AwayTeam=prem2015_16$AwayTeam,FTHG=prem2015_16$FTHG,FTAG=prem2015_16$FTAG)

fulldf<-rbind(prem2015_16,prem2016_17,prem2017_18,prem2018_19,prem2019_20,prem2020_21)
```
Next we find all the teams which have games included in our data set as well as defining $tn$ as the total number of teams.
```{r}
fullteams<-function(dt){
  gamenumber<-length(dt$HomeTeam)
  teams <- vector()
  for (i in 1:gamenumber){
    if (dt$HomeTeam[i] %in% teams==FALSE){
      teams <- append(teams, dt$HomeTeam[i])
      }
  }
  return(teams)
  }
fullteamlist<-sort(fullteams(fulldf))
tn<-length(fullteamlist)

```
Now we define the vector monthd. This contains the number of days in each month.
```{r}
monthd<-c(31,28,31,30,31,30,31,31,30,31,30,31)
```
As our weighting function uses $t$ in units of days back we need to define a function that, using the date a match took place on and the date on which we are predicting each team's strength, can calculate how many days back a match took place.
```{r}
datetodays<-function(date){
  year<-as.numeric(substr(date,7,10))
  month<-as.numeric(substr(date,4,5))
  day<-as.numeric(substr(date,1,2))
  yeardays<-365*(year-1)
  monthdays<-0
  for (i in 1:month){
    monthdays<-monthdays+monthd[i]
  }
  if (year==2020){
    if (month>1){
      monthdays<-monthdays+1
    }
  }
  days<-day+monthdays+yeardays
  return(days)
  }

```
We now convert our dataframe so it only involves the relevant information. That involves converting the date each match took place on into how many days it took place before the data set. This also involves defining the result of each match by comparing how many goals each team scored in each match.
```{r}
converted<-function(dt){
  teams<-fullteams(dt)
  n<-length(teams)
  gn<-length(dt$HomeTeam)
  finaldate<-dt$Date[gn]
  finaldays<-datetodays(finaldate)
  TeamWins<-data.frame(Date=character(0),HomeTeam=character(0),AwayTeam=character(0),Result=integer(0),DaysBack=integer(0))
  for (i in 1:gn){
    Date<-dt$Date[i]
    HomeTeam<-dt$HomeTeam[i]
    AwayTeam<-dt$AwayTeam[i]
    if (dt$FTHG[i]>dt$FTAG[i]){
          Result<-1
          }
        if (dt$FTHG[i]<dt$FTAG[i]){
          Result<-(-1)
          }
        if (dt$FTHG[i]==dt$FTAG[i]){
          Result<-0
        }
    DaysBack<-finaldays-datetodays(dt$Date[i])
    gameresult<-data.frame(Date,HomeTeam, AwayTeam, Result, DaysBack)
    TeamWins<- rbind(TeamWins, gameresult)
    }
  return(TeamWins)
}
dtfrd<-converted(fulldf)
```
As we will need to restimate the statistical parameters every half week to ensure our estimations are up to date we will now code a way to select all the matches before a specific date and also the first five matches after a specific date. This is because, for each round of estimations, we will use all previous data to make these estimations. We then will use these estimations to predict the next five matches as the next five matches account for one half week of matches. 
```{r}
priorgames<-function(df,gwback){
  gn<-length(df$Date)
  countback<-gn+1-(5*gwback)
  beforedf<-df
  beforedf<-beforedf[-(countback:gn), ]
  beforedf$DaysBack<-beforedf$DaysBack - (df$DaysBack[countback])
  return(beforedf)
}

aftergames<-function(df,gwback){
  gn<-length(df$Date)
  countback<-gn-(5*gwback)
  afterdf<-df
  afterdf<-afterdf[((countback+1):(countback+5)), ]
  return(afterdf)
}
```
We now code our log-likelihood function that takes in account home advantage and draws. This is $$\ell(\pi,\gamma,\theta)=\sum_{i}\sum_{j}[(a_{ij}+v_{ij})\ln{(\gamma\pi_i)}-(a_{ij}+v_{ij})\ln{(\gamma\pi_i+\theta\pi_j)}+(b_{ij}+v_{ij})\ln{(\pi_j)}-(b_{ij}+v_{ij})\ln{(\pi_j+\theta\gamma\pi_i)}+v_{ij}\ln{(\theta^2-1)}]$$. We define the strength of the last team to be 0.1. We use $\pi_i=\exp{(b_i)}$ to ensure all our values are positive and logSumExp for its computational advantages. We then use the optim function to find our maximum likelihood estimators. Finally we add in our given strength of the last team to ensure we have a full length vector containing all the statistical parameters.
```{r}
MLE<-function(eps,df){
  dc<-eps
  gn<-length(df$HomeTeam)
  WDHLL3<-function(theta){
    theta[tn+2]<-theta[tn+1]
    theta[tn+1]<-theta[tn]
    theta[tn]<-log(0.1)
    D<-theta[tn+1]
    H<-theta[tn+2]
    ans<-0
    for (i in 1:gn){
      Home<-df$HomeTeam[i]
      Away<-df$AwayTeam[i]
      Result<-df$Result[i]
      Days<-df$DaysBack[i]
      I<-theta[which(fullteamlist==Home)]
      J<-theta[which(fullteamlist==Away)]
      decay<-exp(-(dc*Days))
      if (Result==1){
        parta<-(H+I)-(logSumExp(c(H+I,D+J)))
      }
      if (Result==-1){
        parta<-(J)-(logSumExp(c(J,D+H+I)))
      }
      if (Result==0){
        parta<-(log((exp(2*D))-1))+(H+I+J)-(logSumExp(c(H+I,D+J)))-(logSumExp(c(J,D+H+I)))
      }
      ans<-ans+(decay*parta)
    }
    return(ans)

  }
  qq<-c(rep(1,tn-1),1.5,1.8)
  lowlim<-c(rep(-100,tn-1),1e-5,1e-5)
  strengthDH2<-optim(theta <- qq, WDHLL3, method="L-BFGS-B",lower=lowlim,control=list(fnscale=-1,maxit=1000))
  strengthDH2<-exp(strengthDH2$par)
  strengthDH2[tn+2]<-strengthDH2[tn+1]
  strengthDH2[tn+1]<-strengthDH2[tn]
  strengthDH2[tn]<-0.1
  return(strengthDH2)
}
```
We know code the function "logprobsingle" that finds the value of $$$\kappa(\epsilon)=\sum_{k=1}^{N}(\delta_{k}^{i}\log{(p_{i,j})}+\delta_{k}^{j}\log{(p_{j,i})}+\delta_{k}^{i=j}\log{(p_{i=j}}))$$ for a single match.
```{r}
logprobsingle<-function(True,probabilities){
  if (True==1){
    logpk<-log(probabilities[1])
  }
  if (True==0){
    logpk<-log(probabilities[2])
  }
  if (True==-1){
    logpk<-log(probabilities[3])
  }
  return(logpk)
}
```
We now code the function "predictor" that uses our Bradley-Terry model to find the probability of each possible result.
```{r}
predictor<-function(hteam,ateam,strengths){
  hstr<-strengths[which(fullteamlist==hteam)]#home strength
  astr<-strengths[which(fullteamlist==ateam)]#away strength
  D<-strengths[tn+1]
  H<-strengths[tn+2]
  hwin<-(H*hstr)/((H*hstr)+(D*astr))
  draw<-(((D^2)-1)*H*hstr*astr)/(((H*hstr)+(D*astr))*((astr)+(H*D*hstr)))
  awin<-(astr)/((astr)+(H*D*hstr))
  return(c(hwin,draw,awin))
}
```
We now combine all our functions we have defined. We estimate the statistical parameters for each half week. After this we find the probabilities and predict the results of the matches in the following 5 matches of the data (we predict the results of the matches in each half week). We will then see how successful our predictions are. We do this by seeing how many predictions we got right as well as find the value of $\kappa$ when we use our model. We use $\epsilon=0.0057$ as we have previously found this to be the optimum value of $\epsilon$ that maximises $\kappa$.
```{r}
seasonpredictors<-function(eps,df,gwback){
  logprob<-0
  correctpredictions<-0
  num_games<-0
  possresults<-c(1,0,-1)
  for (i in 1:gwback){
    before<-priorgames(df,i)
    after<-aftergames(df,i)
    MLEs<-MLE(eps,before)
    for (j in 1:5){
      num_games<-num_games+1
      prediction<-predictor(after$HomeTeam[j],after$AwayTeam[j],MLEs)
      logprob<-logprob+logprobsingle(after$Result[j],prediction)
      ourprediction=possresults[which(prediction==max(prediction))]
      if (ourprediction==after$Result[j]){
        correctpredictions<-correctpredictions+1
      }
    }
  }
  return(c(correctpredictions,num_games,logprob))  
}
```

```{r}
seasonpredictors(0.0022,dtfrd,30)
```

