---
title: "draw BT sim"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(matrixStats)
```

First we define the names of each team as well as a function that allows us to select the name of each team as long as we know which number element in the list they occur. We then define $tn$ as the total number of teams.

```{r}
namesa<-c("Ta1","Ta2","Ta3","Ta4","Ta5","Ta6","Ta7","Ta8","a9","Taten","Tb1","Tb2","Tb3","Tb4","Tb5","Tb6","Tb7","Tb8","Tb9","Tbten")  #defining the names for each team
namesb<-c("Ta1","Ta2","Ta3","Ta4","Ta5","Ta6","Ta7","Ta8","a9","Taten","Tb1","Tb2","Tb3","Tb4","Tb5","Tb6","Tb7","Tb8","Tb9","Draw") #defining the names for each team except the final team as well as the name "draw" for later use as well
words<-function(i){   #a function that returns the team we want
  a<-namesa
  return(a[i])
}
tn<-length(namesa) #number of teams
```

Next we define the Bradley-Terry model. We will use the same one for all different versions of the Bradley-Terry model, but with home/draw parameters being 1 if they are not included. In this model we have that the home parameter=1 but the draw parameter does not.

```{r}
BTM<-function(i,j,theta){
  i<-theta[i]#strength of team i
  j<-theta[j]#strength of team j
  d<-theta[tn+1]# 1 if not included
  h<-1# 1 if not included
  iprob<-(i*h)/((h*i)+(d*j)) #prob team i wins following Bradley-Terry model
  jprob<-j/((h*d*i)+j)#prob team j wins following Bradley-Terry model
  tprob<-(((d^2)-1)*h*i*j)/(((h*i)+(d*j))*(j+(d*h*i)))#prob of draw following Bradley-Terry model
  return(c(iprob,jprob,tprob))
}
```
The next function simulates the result of one game between two teams, $i$ and $j$. It does this by running the function BTM that we just defined and getting the probability $i$ wins, probability j wins and probability of a tie. We also simulate a random $Uniform(0,1)$ observation. Call this $x$. If $x$ is less than or equal to probability team $i$ win, then the result is a win for team $i$. If $x$ is more than probability team $i$ wins but less then or equal to the probability of a draw + the probability team $i$ wins then the result is a draw. If $x$ is more than the probability of a draw + the probability team $i$ wins then team $j$ wins. we then record all the results in a dataframe.

```{r}
dgamesimulator<-function(i,j,theta){
  iwin<-0
  jwin<-0
  draw<-0
  HomeTeam <- c(words(i)) #selecting home team
  AwayTeam <-c(words(j)) #selecting away team
  rand<-runif(1) #randomly simulating x
  prob<-BTM(i,j,theta) #finding the probability of each side winning and a draw
  
  #next bit works out what the result of the game is and gives it a value of 1 if that event happens
  if(rand<=prob[1]){
    iwin<-1
  }
  if(prob[1]<rand & rand<(prob[1]+prob[3])){
    draw<-1
  }
  if(rand>(prob[1]+prob[3])){
    jwin<-1
  }
  
  HomeWins <- c(iwin)
  HomeLosses <- c(jwin)
  Ties<-c(draw)
  
  wins.data <- data.frame(HomeTeam, AwayTeam, HomeWins, HomeLosses,Ties) #putting the result of the match in a data frame
  return(wins.data)
  }
```

Next we simulate all the matches in a season. We do this by running 10 home games and 10 away games for each pair of teams using the previously defined function dgamesimulator. We then use rbind to merge the results into one data frame so we have a data frame, dtfr, containing the results of every game in the season.

```{r}
allsim<-function(theta){
  TeamWins<-data.frame(Team1=character(0),Team2=character(0),HomeWins=integer(0),AwayWins=integer(0),Ties=integer(0))
  for (i in 1:tn){ 
    for (j in 1:tn){
      if ((i!=j)){ #as a team cannot play itself
        game1<-dgamesimulator(i,j,theta)
        game2<-dgamesimulator(i,j,theta)
        game3<-dgamesimulator(i,j,theta)
        game4<-dgamesimulator(i,j,theta)
        game5<-dgamesimulator(i,j,theta)
        game6<-dgamesimulator(i,j,theta)
        game7<-dgamesimulator(i,j,theta)
        game8<-dgamesimulator(i,j,theta)
        game9<-dgamesimulator(i,j,theta)
        game10<-dgamesimulator(i,j,theta)
      TeamWins <- rbind(TeamWins, game1,game2,game3,game4,game5,game6,game7,game8,game9,game10)
      }
    }
  }
  return(TeamWins)
}
```

Here we define the true value for theta and then use the function "allsim" to simulate the season of matches and create the dataframe. Thetaa is all the elements in theta except the strength of the last team and is only defined for analysis later.

```{r}
theta<-c(seq(0.05,1,0.05),1.8)
Ttheta<-c(seq(0.05,0.95,0.05),1.8)
dtfr<-allsim(theta)
dtfr
```

The next function counts the number of wins, losses and draws between each team. It does this by frst using the "grep" function to finding all the games in dtfr with the required home team. We then use the "grep" function again on this set of matches to find just the matches with the correct home and away team. We then sum the number of home wins, home losses and ties to find the number of times each of these events occur in matches between the two teams.

```{r}
homefunction<-function(i,j,dtfr){
  if (i==j){
    return(c(0,0,0))#as a team cannot play itself
  }
  hometeamgames<-dtfr[grep(i, dtfr$HomeTeam), ]
  correctgame<-hometeamgames[grep(j, hometeamgames$AwayTeam), ]
  aij<-sum(correctgame$HomeWins)
  bij<-sum(correctgame$HomeLosses)
  vij<-sum(correctgame$Ties)
  return(c(aij,bij,vij))
  
}

```

The following functions find the total number of wins and ties for the two teams, $i$ and $j$, including all the games where $i$ is home and $j$ is away and vice-versa. We do this by running the previously defined function "homefunction" for the pair of teams twice (swapping which one is at home) and adding together the necessary values.

```{r}
w<-function(i,j,dtfr){
  wij<-homefunction(i,j,dtfr)[1]+homefunction(j,i,dtfr)[2]
  return(wij)
}

t<-function(i,j,dtfr){
  tij<-homefunction(i,j,dtfr)[3]+homefunction(j,i,dtfr)[3]
  return(tij)
}
```
Again we define the Bradley-Terry model but this time the Rao-Kupper extension as this takes draws into account. The log likelihood function we have coded is $$\ell(\pi,\theta)=\sum_{i}\sum_{j}[(w_{ij}+t_{ij})(\ln{(\pi_i)}-\ln{(\pi_i+\theta\pi_j)})+\frac{1}{2}t_{ij}\ln{(\theta^2-1)}]$$ and this is what we have coded. We use $\pi_i=\exp{(b_i)}$ to ensure all our values are positive and logSumExp for its computational advantages. We also set the strength of the last team to be the log of its true strength to make our results comparable. We then use a pair of "for" functions for $i$ in $1:tn$ and $j$ in $1:tn$ to account for all teams and all values of $i$ and $j$.

```{r}
RKLL2<-function(theta){
  theta[tn+1]<-theta[tn]
  theta[tn]<-log(1)
  D<-(theta[tn+1])
  ans<-0
  for (i in 1:tn){
    teami<-words(i)
    I<-(theta[i])
    for (j in 1:tn){
      teamj<-words(j)
      J<-(theta[j])
      wij<-w(teami,teamj,dtfr)
      tij<-t(teami,teamj,dtfr)
      parta<-(wij+tij)*(I-logSumExp(c(I,D+J)))
      partb<-(1/2)*tij*log(exp(2*D)-1)
      ans<-ans+parta+partb
      
        
    }
  }
  return(ans)
}

```  

Next we find the MLEs using the optim function. We use fnscale=-1 to find the maximum. The lower limit for the draw parameter is set approximately to zero. This is because we need $\theta>1$ which is equvalent to $ln(\theta)>0$.

```{r}
a<-c(rep(0.1,19),1.5)
lowlim<-c(rep(-100,19),1e-5)
strengthRK<-optim(theta <- a, RKLL2, method="L-BFGS-B",lower=lowlim,control=list(fnscale=-1),hessian=TRUE)
strengthRK
```

We now use the hessian matrix the calculate the fisher-information matrix. Then, using this, we calculate the corresponding confidence-intervals for each statistical parameter and assess whether the true value for each statistical parameter is contained within each confidence-interval.

```{r}
fisher_info<-solve(-strengthRK$hessian)
prop_sigma<-sqrt(diag(fisher_info))
upper<-strengthRK$par+1.96*prop_sigma
lower<-strengthRK$par-1.96*prop_sigma
interval<-(data.frame(Team=namesb,value=exp(strengthRK$par), upper=exp(upper), lower=exp(lower), True=Ttheta))
interval
```







