---
title: "original BT sim"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(matrixStats)
```

First we define the names of each team as well as a function that allows us to select the name of each team as long as we know which number element in the list they occur. We then define $tn$ as the total number of teams.

```{r}
namesa<-c("Ta1","Ta2","Ta3","Ta4","Ta5","Ta6","Ta7","Ta8","a9","Taten","Tb1","Tb2","Tb3","Tb4","Tb5","Tb6","Tb7","Tb8","Tb9","Tbten")
namesb<-c("Ta1","Ta2","Ta3","Ta4","Ta5","Ta6","Ta7","Ta8","a9","Taten","Tb1","Tb2","Tb3","Tb4","Tb5","Tb6","Tb7","Tb8","Tb9")#defining the names for each team
words<-function(i){   #a function that returns the team we want
  a<-namesa
  return(a[i])
}
tn<-length(namesa) #number of teams
```

Next we define Bradley Terry model. we will use the same one for all different bradley terry models but with home/draw parameters being 1 if they are not included. we can see this is ok from a quick observation of the given Bradley Terry Models.

```{r}
BTM<-function(i,j,theta){
  i<-theta[i]#strength of team i
  j<-theta[j]#strength of team j
  d<-1# 1 if not included
  h<-1# 1 if not included
  iprob<-(i*h)/((h*i)+(d*j)) #prob team i wins following Bradley-Terry model
  jprob<-j/((h*d*i)+j)#prob team j wins following Bradley-Terry model
  tprob<-(((d^2)-1)*h*i*j)/(((h*i)+(d*j))*(j+(d*h*i)))#prob of draw following bradley terry model
  return(c(iprob,jprob,tprob))
}
```

The next function simulates the result of one game between the two teams. It does this by running the function BTM that we just defined and getting the prob i wins, prob j wins and prob of a tie. We also simulate a random U(0,1) observation. Call this x. If x is less than or equal to probi then the result is team i wins. If x is more than prob team i wins but less then or equal to (prob of a draw + prob team i wins) then the result is a draw. If x is more than (prob of a draw + prob team i wins) then team j wins. we then record all the results in a dataframe.

```{r}
dgamesimulator<-function(i,j,theta){
  iwin<-0
  jwin<-0
  draw<-0
  HomeTeam <- c(words(i)) #selecting home team
  AwayTeam <-c(words(j)) #selecting away team
  rand<-runif(1) #randomly simulating x
  prob<-BTM(i,j,theta) #finding the probability of each side winning and a draw
  
  #next bit works out what the result of the game is and gives it a value of 1 if that event happens
  if(rand<=prob[1]){
    iwin<-1
  }
  if(prob[1]<rand & rand<(prob[1]+prob[3])){
    draw<-1
  }
  if(rand>(prob[1]+prob[3])){
    jwin<-1
  }
  
  HomeWins <- c(iwin)
  HomeLosses <- c(jwin)
  Ties<-c(draw)
  
  wins.data <- data.frame(HomeTeam, AwayTeam, HomeWins, HomeLosses,Ties) #putting the result of the match in a data frame
  return(wins.data)
  }
```

Next we simulate all the matches in a season. We do this by running 8 home games and 8 away games for each pair of teams using the previously defined function dgamesimulator. We then use rbind to merge the results into one data frame so we have a data frame, dtfr, containing the results of every game in the season.

```{r}
allsim<-function(theta){
  TeamWins<-data.frame(Team1=character(0),Team2=character(0),HomeWins=integer(0),AwayWins=integer(0),Ties=integer(0))
  for (i in 1:tn){ 
    for (j in 1:tn){
      if ((i!=j)){ #as a team cannot play itself
        game1<-dgamesimulator(i,j,theta)
        game2<-dgamesimulator(i,j,theta)
        game3<-dgamesimulator(i,j,theta)
        game4<-dgamesimulator(i,j,theta)
        game5<-dgamesimulator(i,j,theta)
        game6<-dgamesimulator(i,j,theta)
        game7<-dgamesimulator(i,j,theta)
        game8<-dgamesimulator(i,j,theta)
        game9<-dgamesimulator(i,j,theta)
        game10<-dgamesimulator(i,j,theta)
        
      TeamWins <- rbind(TeamWins, game1,game2,game3,game4,game5,game6,game7,game8,game9,game10)
      }
    }
  }
  return(TeamWins)
}
```

Here we define the true value for theta and then use the function "allsim" to simulate the season of matches and create the dataframe. Thetaa is all the elements in theta except the last one and is only defined for analysis later.

```{r}
theta<-c(0.05,0.1,0.15,0.2,0.25,0.3,0.35,0.4,0.45,0.5,0.55,0.6,0.65,0.7,0.75,0.8,0.85,0.9,0.95,1.0)
Ttheta<-c(0.05,0.1,0.15,0.2,0.25,0.3,0.35,0.4,0.45,0.5,0.55,0.6,0.65,0.7,0.75,0.8,0.85,0.9,0.95)

dtfr<-allsim(theta)
dtfr
```

The next function counts the number of wins, losses and draws between each team. It does this by frst using the "grep" function to finding all the games in dtfr with the required home team. We then use the "grep" function again on this set of matches to find just the matches with the correct home and away team. We then sum the number of home wins, home losses and ties to find the number of times each of these events occur in matches between the two teams.

```{r}
homefunction<-function(i,j,dtfr){
  if (i==j){
    return(c(0,0,0))#as a team cannot play itself
  }
  hometeamgames<-dtfr[grep(i, dtfr$HomeTeam), ]
  correctgame<-hometeamgames[grep(j, hometeamgames$AwayTeam), ]
  aij<-sum(correctgame$HomeWins)
  bij<-sum(correctgame$HomeLosses)
  vij<-sum(correctgame$Ties)
  return(c(aij,bij,vij))
  
}
```

The following functions find the total number of wins and ties for the two teams, a and b, including all the games where a is home and b is away and vice versa. We do this by running the previously defined function "homefunction" for the pair of teams twice (swapping which one is at home) and adding together the neccessary values

```{r}
w<-function(i,j,dtfr){
  wij<-homefunction(i,j,dtfr)[1]+homefunction(j,i,dtfr)[2]
  return(wij)
}

t<-function(i,j,dtfr){
  tij<-homefunction(i,j,dtfr)[3]+homefunction(j,i,dtfr)[3]
  return(tij)
}
```

```{r}
sum(dtfr$HomeWins)
sum(dtfr$HomeLosses)


```

Now we define our log likelihood function for the first general Bradley Terry model following the log-likelihood function, $$\ell(\pi) =  \sum_{i=1}^{\infty}\sum_{j=1}^{\infty} [w_{ij}\ln{(\pi_i)} - w_{ij}\ln{(\pi_i+\pi_j)}]$$ . we use pi=exp(bi) and logSumExp for its numerical advantages. the sum to i of the sum to j is taken into account by starting off with a value of zero. We then use a pair of "for" functions for i in 1:10 and j in 1:10 to account for all teams and all values of i and j

```{r}
BTLL<-function(theta){
  theta[tn]<-log(1)
  ans = 0
  for (i in 1:tn){
    teami<-words(i)
    for (j in 1:tn){
      teamj<-words(j)
      wij<-w(teami,teamj,dtfr)
      ans = ans + ((wij*log(exp(theta[i])))-((wij)*logSumExp(c(theta[i],theta[j]))))
      }
    }
  return(ans)
}
```

Next we find the MLEs using the optim function. We use fnscale=-1 to find the maximum.

```{r}
a<-rep(0.1,19)
strength<-optim(theta <- a, BTLL, method="L-BFGS-B",control=list(fnscale=-1),hessian=TRUE)
strength
```
We now use the hessian matrix the calculate the fisher-information matrix. Then, using this, we calculate the corresponding confidence-intervals for each statistical parameter and assess whether the true value for each statistical parameter is contained within each confidence-interval.
```{r}
fisher_info<-solve(-strength$hessian)
prop_sigma<-sqrt(diag(fisher_info))
upper<-strength$par+1.96*prop_sigma
lower<-strength$par-1.96*prop_sigma
interval<-(data.frame(Team=namesb,value=exp(strength$par), upper=exp(upper), lower=exp(lower), True=Ttheta))
interval
```











